The Birth of a Galaxy: Population III Progenitors and Metal-enriched Stellar Populations

### 2010-10-12 ms.tex
Added figures 3 and 4.  Updated figures 1 and 2 to z=7.
First draft of first results subsection, halo evolution.

The mass-dependent luminosities and lifetimes of the Pop III stars are
taken from \citet{Schaerer02}.  The radiation field is evolved with
with adaptive ray tracing \citep{Abel02_RT, Wise10} and is coupled
self-consistently to the hydrodynamics.  These stars die as
pair-instability SNe (PISNe) if they are in the mass range 140--260
\Ms \citep{Heger03}.  We use the explosion energy from
\citet{Heger02}, where we fit the following function to their models,
$E_{\rm PISN} = 10^{51} \times [5.0 + 1.304 (M_{\rm He} - 64)]$, where
$M_{\rm He} = (13/24) \times (M_\star - 20) \Ms$ is the helium core
mass and $M_\star$ is the stellar mass.

@@

Here we present the gaseous and stellar evolution of three selected
halos in the simulation: (1) a halo that never recovers a gas
reservior after a PISN, (2) one that experiences a series of major
mergers, and (3) one that has an early mass buildup but no major
mergers after $z=12$.  The halos are referred to by these numbers.  We
start with the birth of the first Pop III star in its merger history.
We then compare the nature of star formation in these halos.

\li Should we present the gas-poor halo?

@@

\begin{figure}
  \epsscale{1.15}
  \plotone{f1}
  \caption{\label{fig:evo}}
  \epsscale{1}
\end{figure}

\begin{figure*}
  \plotone{f2}
  \caption{\label{fig:projections}}
\end{figure*}

\li Describe the evolution of the baryon fraction and metallicity from
PISN and Type II SNe metals in the two halos.

\li Need to run the halo analysis on smaller halos to capture the
entire merger histories.

@@

\subsection{Population II stellar populations}

@@

Support for this work was provided by NASA through Hubble Fellowship
grant \#120-6370 awarded by the Space Telescope Science Institute,
which is operated by the Association of Universities for Research in
Astronomy, Inc., for NASA, under contract NAS 5-26555.  Computational
resources were provided by NASA/NCCS award SMD-09-1439.  The majority
of the analysis and plots were done with \texttt{yt} \citep{yt}.

>>>>

The mass-dependent luminosities and lifetimes of the Pop III stars are
taken from \citet{Schaerer02}.  The radiation field is evolved with
with adaptive ray tracing \citep{Abel02_RT, Wise10} and is coupled
self-consistently to the hydrodynamics.  We model the \hh~dissociating
radiation with an optically-thin, inverse square profile, centered on
all stars.  These stars die as pair-instability SNe (PISNe) if they
are in the mass range 140--260 \Ms \citep{Heger03}.  We use the
explosion energy from \citet{Heger02}, where we fit the following
function to their models, $E_{\rm PISN} = 10^{51} \times [5.0 + 1.304
(M_{\rm He} - 64)]$, where $M_{\rm He} = (13/24) \times (M_\star - 20)
\Ms$ is the helium core mass and $M_\star$ is the stellar mass.

@@

\begin{figure*}
  \plotone{f3}
  \caption{\label{fig:projections} Density-weighted projections of gas
    density (top), temperature (middle), and metallicity (bottom).
    The left column shows the entire simulation volume, where the
    center and right columns focus on the intense and quiet halos,
    respectively.  The metallicity projections are a composite picture
    of metals originating from Pop III (red) and Pop II (blue) stars.}
\end{figure*}

Here we present the gaseous and stellar evolution of two selected
halos in the simulation: one that has an early mass buildup but no
major mergers after $z=12$, and one that experiences a series of major
mergers between $z=10$ and $z=7$.  We name the halos ``quiet'' and
``intense'', respectively.  We start with the birth of the first Pop
III star in its merger history.  We then compare the nature of star
formation in these halos.

We illustrate the state of the simulation at $z=7$ in Figure
\ref{fig:projections} with density weighted projections of gas
density, temperature, and metallicity, showing the entire box and
focusing on the two halos of interest.  The quiet and intense halos
are located in the bottom center and bottom right of the full box,
respectively.

@@

\begin{figure}
  \epsscale{1.15}
  \plotone{f1}
  \caption{\label{fig:evo} Evolution of the total halo mass (top),
    stellar mass (middle), and gas fraction (bottom) of the quiet
    (dashed) and intense (solid) halos.}
  \epsscale{1}
\end{figure}

\begin{figure}
  \epsscale{1.15}
  \plotone{f2}
  \caption{\label{fig:Zevo} Average stellar metallicities and gas
    metallicities enriched by Pop II and Pop III SNe of the
    intense (top) and quiet (bottom) halos.}
  \epsscale{1}
\end{figure}

%\li Describe the evolution of the baryon fraction and metallicity from
%PISN and Type II SNe metals in the two halos.

Figure \ref{fig:evo} shows the total, metal-enriched stellar, and gas
mass history of the most massive progenitors of both halos.  The quiet
halo undergoes a series of major mergers at $z > 12$, growing by a
factor of 30 to $2.5 \times 10^7 \Ms$ within 150 Myr.  Afterwards it
only grows by a factor of 3 by $z=7$ mainly through smooth accretion
from the filaments and IGM.  It is the most massive halo in the
simulation between redshifts 13 and 10.  At the same time, the intense
halo has a relatively small mass $M = 10^7 \Ms$, but it is contained
in a biased region on a comoving scale of 50 kpc with $\sim30$ halos
with $M > 10^6 \Ms$.  After $z=10$, these halos hierarchically merge
to form a $10^9 \Ms$ halo at $z=7$ with two major mergers at redshifts
10 and 7.9, seen in the rapid increases in total mass in Figure
\ref{fig:evo}.  The merger history of the two halos are not atypical
as dark matter halos can experience both quiescent and vigorous mass
accretion rates in their buildup.

Both halos star forming Pop II stars when $M = 10^7 \Ms$.  This is
consistent with the filtering mass $M_f$ of high-redshift halos when
it accretes mainly from a pre-heated IGM \citep{gnedin98, gnedin00,
  Wise08_Gal}.  The quiet halo forms $10^5 \Ms$ of stars by $z=9$.
This initial starburst photo-evaporates the majority of its molecular
clouds, in addition to heating and ionizing the surrounding IGM out to
a radius of 10--15 kpc at $z=9$.  These respectively reduce the
in-situ and external cold gas supply that could feed future star
formation.

The gas fractions of both halos are depleted from 0.15 to 0.08 by
outflowing shocks driven by ionization fronts in their initial
starbursts.  The quiet halo never recovers to the cosmic fraction
$\Omega_b/\Omega_M$ for the following reason.  The lack of major
mergers of halos above the filtering mass leads to a small gas
fraction, cold gas reservoir, and stellar mass at later times.  This
plays a large role in gas accretion because halos with $M < M_f$ will
be likely photo-evaporated, hosting diffuse warm gas reservoirs
instead of cold dense cores.  During this quienscent period, the halo
mainly accretes warm diffuse gas from the filaments and IGM.  In
contrast, the intense halo grows from major mergers of halos with $M >
M_f$.  The progenitor halos involved in the major mergers are able to
host molecular clouds and have, in general, higher gas fractions.
Between $z=10$ and $z=8$, the gas fraction increases from 0.07 to 0.12
until it rapidly jumps to 0.14 when a gas-rich major merger occurs.
The stellar mass accordingly increases with the ample supply of cold
gas during this period.  The gas fraction then decreases slightly by
0.01 when galactic outflows are generated from the starburst
associated with this major merger.

The evolution of the stellar and gas metallicity of both halos are
illustrated in Figure \ref{fig:Zevo}.  PISNe from Pop III stars enrich
the nearby IGM out to a radius of 10 kpc and provides a metallicity
floor of $[Z_3/H] \sim -3$ in both halos, given $M_{\rm char} = 100
\Ms$.  The metallicity from Pop II SNe initially enrich the ISM of
both halos to an average $[\bar{Z}_2/H]$ between --3.5 and --3.  The
quiet halo is enriched to $[\bar{Z}_2/H] \sim -2.5$ until the
metal-poor inflows balance the metal-rich galactic outflows, which are
directed in the polar directions of the gas disk.  This is apparent in
the metallicity projections in Figure \ref{fig:projections}.  The
average stellar metallicities is equal or slightly higher up to 0.5
dex than $[\bar{Z}_2/H]$.

At $z \sim 8$, the metallicity of the intense halo dramatically
self-enriches itself by a factor of 30 to $[\bar{Z}_2/H] = 1.5$ during
a starburst.  Because this halo is located in a large-scale
overdensity, the ejecta falls back into the halo after reaching
distances up to 20 comoving kpc, in turn keeping the halo gas
metallicity high because the inflows are relatively metal-rich
themselves.  Interestingly the first few Pop II star clusters have
[Z/H] between --1 and --2 and dominate the average stellar metallicity
at $z > 8$.  We further detail the formation of these clusters in
\S\ref{sec:pop}.  After the $z=8$ starburst, the average stellar
metallicity follows the average gas metallicity within 0.1 dex.

@@

The most massive progenitor of the quiet halo interestingly never
hosts a Pop III star.  Instead a nearby halo forms a Pop III star,
which (randomly) produces a PISNe at $z=16$.  The associated blastwave
overruns the most massive progenitor, and the dense core survives this
event and is enriched by this PISN.

\subsection{Population II stellar populations}
\label{sec:pop}

\begin{figure*}
  \plottwo{f4a}{f4b}
  \caption{\label{fig:pops} The scatter plots show the star formation
    history of the intense (left) and quiet (right) halos as a
    function of metallicity at $z=7$.  Each circle represents a star
    cluster, whose area is proportional to its mass.  The open circles
    in the upper right represent sizes of $10^3$, $10^4$, and $10^5$
    \Ms~star clusters.  The upper histogram collects the star formation
    history into 25 equal temporal bins.  The right histogram depicts
    the stellar metallicity distribution.}
\end{figure*}

@@

Support for this work was provided by NASA through Hubble Fellowship
grant \#120-6370 awarded by the Space Telescope Science Institute,
which is operated by the Association of Universities for Research in
Astronomy, Inc., for NASA, under contract NAS 5-26555.  Computational
resources were provided by NASA/NCCS award SMD-09-1439.  The majority
of the analysis and plots were done with \texttt{yt}
\citep{yt_full_paper}.
